State-Space Model to Estimate Salmon Escapement Using Multiple Data Sources

### 2020-09-17 analysis/paper/appendices/Appendix_STADEM_descrip.Rmd
corrected some references in the appendices, after updating the reference.bib file

Escapement at Lower Granite Dam (LGD) is estimated by combining two independent observations (trap catches and window counts) of the true number of fish crossing the dam in a state-space model [@royle2008hierarchical] with a weekly time-step. Both are assumed to be corrupted observations of the true unknown number of fish crossing LGD each week. The log of the true number of fish crossing ($X_t$), is modeled as a random walk process [@Shumway2010].

>>>>

Escapement at Lower Granite Dam (LGD) is estimated by combining two independent observations (trap catches and window counts) of the true number of fish crossing the dam in a state-space model [@Royle2008] with a weekly time-step. Both are assumed to be corrupted observations of the true unknown number of fish crossing LGD each week. The log of the true number of fish crossing ($X_t$), is modeled as a random walk process [@Shumway2010].

### 2020-09-17 analysis/paper/appendices/Appendix_simulation_details.Rmd
made the equations using normal distributation match the notation in the other appendix
added some \left( and \right) to some latex code.
added a sentence at the end pointing to the github repo for all the code.
corrected some references in the appendices, after updating the reference.bib file

$$
\begin{aligned}
  \zeta_{j,p} &\sim \text{Bernoulli}(\tau_j) \\
  \left[\phi_{j,p=1},  ... , \phi_{j,p=n}\right] &\sim \text{Dir}(\zeta_{j,p=1},  ... , \zeta_{j,p=n}) \\
  N_{j,p} &= N_j * \phi_{j,p} \\
  N_p &= \sum_{j=w, h, hnc} N_{j,p}
\end{aligned}
$$

@@

$$
\begin{aligned}
 \left[\bar{a}_p, ..., \bar{a}_n\right] &\sim \text{Norm}(\mu_a, \sigma^2_a) \\
 \left[s_p, ..., s_n\right] &\sim \left | \text{Norm}(\mu_s, \sigma^2_s) \right |
\end{aligned}
$$

@@

$$
\begin{aligned}
 date_{i,p} &\sim \text{Norm}(\bar{a}_p, s^2_p) \\
\end{aligned}
$$

@@

For the specific simulated scenarios, we were interested in STADEM model estimates of origin specific escapement from the combinations of two separate trapping rates, two fallback, re-ascension and night-passage combinations and three window count error rates; resulting in twelve different scenarios. First, trapping rates were set static at 0.15 across all weeks for six scenarios to mimic an optimum trap operation for an expected return of 25,000 wild fish (i.e., trap $\approx$ 4,000 wild fish). For the remaining six scenarios, trapping rates for weeks 30, 31 and 32 (i.e., July $22^{nd}$ to August $11^{th}$) were changed to 0.00 to test STADEM sensitivities to potential trap shut downs similar to those observed in 2013, 2014 and 2015 [@Ogden2014; @Ogden2016; @Ogden2016b]. To simulate and control the number of re-ascending and night-time passing fish to model response, we altered fallback and night-time passage rates while holding the re-ascension rate constant at $\rho = 1.0$. Altering fallback rates and holding re-ascension constant allowed for a more simple control of the number of fish re-ascending; because the number of re-ascending fish is a function of the number of fallbacks and the re-ascension rate. Six scenarios had equal rates of fallback and night-time passage set at $\psi = \nu = 0.06$ [@Boggs2004] which means other estimator assumptions [@Schrader2013]. The other six scenarios set fallback at $\psi = 0.10$ and night-time passage at $\nu = 0.05$ to create a 5.0% positive bias of unique fish at the window. A potential 5.0% weekly bias was determined from PIT-tag data and within the range of observed weekly difference for return years 2010 - 2015 (Figure \@ref(fig:night-reasc-diff-fig)). Finally, we desired to test the sensitivities of STADEM to potential rates of window count error; 0%, 5% and 10% [@Hatch1994]. To simulate window count error, we assumed the observed daily count was a random variable from a normal distribution with a mean equal to the true daily count, and a standard deviation equal to the applied error rate (i.e., 0%, 5%, 10%) multiplied by the true daily count. This method simulated observed counts as unbiased, and allowed for possible under and overcounts at the window.

>>>>

$$
\begin{aligned}
  \zeta_{j,p} &\sim \text{Bernoulli}(\tau_j) \\
  \left[\phi_{j,p=1},  ... , \phi_{j,p=n}\right] &\sim \text{Dir} \left(\zeta_{j,p=1},  ... , \zeta_{j,p=n} \right) \\
  N_{j,p} &= N_j * \phi_{j,p} \\
  N_p &= \sum_{j \in w, h, hnc} N_{j,p}
\end{aligned}
$$

@@

$$
\begin{aligned}
 \left[\bar{a}_p, ..., \bar{a}_n\right] &\sim \mathcal{N}(\mu_a, \sigma^2_a) \\
 \left[s_p, ..., s_n\right] &\sim \left | \mathcal{N}(\mu_s, \sigma^2_s) \right |
\end{aligned}
$$

@@

$$
\begin{aligned}
 date_{i,p} &\sim \mathcal{N}(\bar{a}_p, s^2_p) \\
\end{aligned}
$$

@@

For the specific simulated scenarios, we were interested in STADEM model estimates of origin specific escapement from the combinations of two separate trapping rates, two fallback, re-ascension and night-passage combinations and three window count error rates; resulting in twelve different scenarios. First, trapping rates were set static at 0.15 across all weeks for six scenarios to mimic an optimum trap operation for an expected return of 25,000 wild fish (i.e., trap $\approx$ 4,000 wild fish). For the remaining six scenarios, trapping rates for weeks 30, 31 and 32 (i.e., July $22^{nd}$ to August $11^{th}$) were changed to 0.00 to test STADEM sensitivities to potential trap shut downs similar to those observed in 2013, 2014 and 2015 [@Ogden2014; @Ogden2016a; @Ogden2016b]. To simulate and control the number of re-ascending and night-time passing fish to model response, we altered fallback and night-time passage rates while holding the re-ascension rate constant at $\rho = 1.0$. Altering fallback rates and holding re-ascension constant allowed for a more simple control of the number of fish re-ascending; because the number of re-ascending fish is a function of the number of fallbacks and the re-ascension rate. Six scenarios had equal rates of fallback and night-time passage set at $\psi = \nu = 0.06$ [@Boggs2004] which means other estimator assumptions [@Schrader2013]. The other six scenarios set fallback at $\psi = 0.10$ and night-time passage at $\nu = 0.05$ to create a 5.0% positive bias of unique fish at the window. A potential 5.0% weekly bias was determined from PIT-tag data and within the range of observed weekly difference for return years 2010 - 2015 (Figure \@ref(fig:night-reasc-diff-fig)). Finally, we desired to test the sensitivities of STADEM to potential rates of window count error; 0%, 5% and 10% [@Hatch1994]. To simulate window count error, we assumed the observed daily count was a random variable from a normal distribution with a mean equal to the true daily count, and a standard deviation equal to the applied error rate (i.e., 0%, 5%, 10%) multiplied by the true daily count. This method simulated observed counts as unbiased, and allowed for possible under and overcounts at the window.

All code for simulating data and fitting STADEM to that data can be found at [https://www.github.com/KevinSee/ManuscriptSTADEM]([https://www.github.com/KevinSee/ManuscriptSTADEM]).
